Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-03-01 NovelLBM.tex
3D shear

@@

\usepackage{bbm} % provides unit tensor via \mathbbm{1}
\newcommand{\ma}[1]{\underline{#1}}
\newcommand{\uT}{\mathbbm{1}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\dif}{d}
\DeclareMathOperator{\Dif}{D}

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
Specifically, we implement the LBM collision as follows.
\begin{equation}
	f_i\left(\vec{x}+\vec{c_i}\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right) + B_i\right]
\end{equation}
Where $f_i$ are the populations, $f_i^\text{eq}$ are the equilibrium populations, $\vec{c_i}$ are the lattice velocities, $\Delta t$ is the time step, $\tau_\text{r}$ is the relaxation time and $B_i$ are the stress contributions defined as follows.
\begin{equation}
	B_i = w_i\left(\frac{\vec{c_i}\vec{c_i}}{2c_\text{s}^4}-\frac{\uT}{2c_\text{s}^2}\right):\ma{\tau}
\end{equation}
Where $w_i$ are the lattice weights, $c_\text{s}$ is the lattice speed of sound and $\ma{\tau}$ is the polymer stress tensor. The colon denotes the Frobenius inner product.
\SG{ more details }\RK{done}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is \RK{hier könnte ein Adjektiv stehen} to the one in Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow, we assume that the velocity is unidirectional:
\begin{equation}
	\vec{u} = u_x\left(y\right)\hat{e}_x
\end{equation}
\begin{equation}
	\dot{\gamma}\left(y\right) = \frac{\partial u_x}{\partial y}
\end{equation}

With this the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state reduces to
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xy} \label{eq:PTTDerivation1}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\dot{\gamma} \label{eq:PTTDerivation2}
for a planar flow.
In axisymmetric flows $y$ in eqs.~(\ref{eq:PTTDerivation1}, \ref{eq:PTTDerivation2}) needs to be replaced by the radial coordinate $r$. \RK{2D case is also axisymmetric?}\SG{ no factor 2? }\RK{No, factor 2 comes from the NS eq. and thus is only relevant for Poiseuille solution. This replacement is also much more relevant there as here all ys disappear}
With $\Tr\ma{\tau} = \tau_{xx}$.
Eliminating $\tau_{xy}$ yields the following.
	\exp\left(2\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} = 2\lambda\dot{\gamma}^2\eta_\text{p}
This is solved by the Lambert $W$ function as follows.\todo{split 2D and 3D derivation}
	\tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)\label{eq:tau_xx}
With $\tau_{yy} = 0$ this yields the first normal stress difference $N_1$ as follows. The second normal stress difference is proportional to $\xi$ and therefore $0$.
	N_1 = \sigma_{xx} - \sigma_{yy} = \tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) %\label{eq:shearN}
For the viscosity, we insert this into equation~(\ref{eq:PTTDerivation2}) and retrieve the following.
	\eta &= \frac{\sigma_{xy}}{\dot{\gamma}} \\
	&= \frac{\tau_{xy} + \eta_\text{s}\dot{\gamma}}{\dot{\gamma}} \\
	&= \eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} %\label{eq:shearEta}
This gives the relevant solutions for the shear flow. For Poiseuille flow, we start with the Cauchy momentum equation (neglecting gravity), which reads as follows.

@@

		r = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}\label{eq:semiAnalyticGD}

>>>>

\usepackage{mleftright}
\mleftright

@@

\usepackage{bbm} % provides unit tensor via \mathbbm{1}
\newcommand{\ma}[1]{\underline{#1}}
\newcommand{\uT}{\mathbbm{1}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\dif}{d}
\DeclareMathOperator{\Dif}{D}
\DeclareMathOperator{\sgn}{sgn}

@@

To include the polymer stress into the LBM, we use the stress coupling scheme according to Dzanic \textit{et al.} \cite{dzanic2022_2, gupta2015, onishi2006}.
Specifically, we implement the LBM collision as follows.
\begin{equation}
	f_i\left(\vec{x}+\vec{c}_i\Delta t, t+\Delta t\right) - f_i\left(\vec{x}, t\right) = -\frac{\Delta t}{\tau_\text{r}}\left[f_i\left(\vec{x}, t\right) - f_i^\text{eq}\left(\vec{x}, t\right) + B_i\right]
\end{equation}
Where $f_i$ are the populations, $f_i^\text{eq}$ are the equilibrium populations, $\vec{c_i}$ are the lattice velocities, $\Delta t$ is the time step, $\tau_\text{r}$ is the relaxation time and $B_i$ are the stress contributions defined as follows.
\begin{equation}
	B_i = w_i\left(\frac{\vec{c}_i\vec{c}_i}{2c_\text{s}^4}-\frac{\uT}{2c_\text{s}^2}\right):\ma{\tau}
\end{equation}
Where $w_i$ are the lattice weights, $c_\text{s}$ is the lattice speed of sound and $\ma{\tau}$ is the polymer stress tensor. The colon denotes the Frobenius inner product.
\SG{ more details }\RK{done}
To handle inflow/outflow boundaries, the polymer conformation tensor is specified on the boundary and advected using the corner transport upwind scheme.

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is similar to the one in Oliveira \cite{oliveira1999}. The origin is placed in the center of the channel. The $x$-axis points in flow direction. For the two-dimensional case the $y$-axis points along the velocity gradient.
We separate the derivation for two and three dimensions for mathematical clarity. They are however similar. We start with shear flow in the two-dimensional case.\todo{maybe}
\subsection{Shear flow}
For both Poiseuille flow and shear flow, symmetry dictates that the velocity is unidirectional:
\begin{equation}
	\vec{u} = u_x\left(r\right)\hat{e}_x
\end{equation}
\begin{equation}
	\dot{\gamma}\left(r\right) = \frac{\partial u_x}{\partial r} = \sgn\left(y\right)\frac{\partial u_x}{\partial y} \leq 0
\end{equation}
With $r = \abs{y}$.
For pure shear flow, $\dot{\gamma}$ is a constant.

With this the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state reduces to the following. Note, that $\ma{\tau}$ is symmetric.\RK{Am I allowed to assume this is trivial?}
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\sgn\left(y\right)\dot{\gamma}\tau_{xy}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\sgn\left(y\right)\dot{\gamma}
\end{align}
With $\Tr\ma{\tau} = \tau_{xx}$. Note that all other components aside from $\tau_{xx}$ and $\tau_{xy}$/$\tau_{yx}$ are zero.
We define $\tau_{xr} = \tau_{xy}\sgn\left(y\right) < 0$ to get the following.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xr} \label{eq:PTTDerivation1}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xr} &= \eta_\text{p}\dot{\gamma} \label{eq:PTTDerivation2}
\end{align}
Eliminating $\tau_{xr}$ yields the following.
\begin{equation}
	\exp\left(2\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} = 2\lambda\dot{\gamma}^2\eta_\text{p}
\end{equation}
This is solved by the Lambert $W$ function as follows.
\begin{equation}
	\tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)\label{eq:tau_xx}
\end{equation}
With $\tau_{yy} = 0$ this yields the first normal stress difference $N_1$ as follows. The second normal stress difference is proportional to $\xi$ and therefore $0$.
\begin{equation}
	N_1 = \sigma_{xx} - \sigma_{yy} = \tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon\mathrm{Wi}^2\right) %\label{eq:shearN}
\end{equation}
For the viscosity, we insert this into equation~(\ref{eq:PTTDerivation2}) and retrieve the following.
\begin{align}
	\eta &= \frac{\sigma_{xr}}{\dot{\gamma}} \\
	&= \frac{\tau_{xr} + \eta_\text{s}\dot{\gamma}}{\dot{\gamma}} \\
	&= \eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} %\label{eq:shearEta}
\end{align}
With $\sigma_{xr} = \sigma_{xy}\sgn\left(y\right) < 0$.
This gives the relevant solutions for shear flow. However, these equations also hold for Poiseuille flow.

In three dimensions we pick cylindrical coordinates as follows.
\begin{align}
	x &= x\\
	y &= r\cos\varphi\\
	z &= r\sin\varphi
\end{align}
Limiting $\varphi$ to $\varphi\in\left\{0, \pi\right\}$ reproduces the above definition of $r$ in two dimensions. Other than this expansion of the definitions, the equations from above are the same in three dimension as they are in two.
\RK{The definitions of $r$ above are entirely valid in 2D and picked in a way to make the signs consistent with 3D. In my opinion this means, that we do not need to show the calculation here. But I could if you want multiple pages of boring math. See:}

\RK{

\begin{equation}
	\nabla\vec{u} = \frac{\partial u_x}{\partial r}\hat{e}_r\otimes\hat{e}_x
\end{equation}
\begin{equation}
	\ma{\tau} = \tau_{ij}\hat{e}_i\otimes\hat{e}_j
\end{equation}
\begin{equation}
\begin{split}
	\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{\tau} = \frac{\partial u_x}{\partial r}\hat{e}_x\otimes\hat{e}_r&\left(\tau_{xx}\hat{e}_x\otimes\hat{e}_x+\tau_{xr}\hat{e}_x\otimes\hat{e}_r+\tau_{x\varphi}\hat{e}_x\otimes\hat{e}_\varphi\right.\\
	&+\tau_{rx}\hat{e}_r\otimes\hat{e}_x+\tau_{rr}\hat{e}_r\otimes\hat{e}_r+\tau_{r\varphi}\hat{e}_r\otimes\hat{e}_\varphi\\
	&\left.{ }+\tau_{\varphi x}\hat{e}_\varphi\otimes\hat{e}_x+\tau_{\varphi r}\hat{e}_\varphi\otimes\hat{e}_r+\tau_{\varphi\varphi}\hat{e}_\varphi\otimes\hat{e}_\varphi\right)
\end{split}
\end{equation}
\begin{equation}
	\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{\tau} = \frac{\partial u_x}{\partial r}\left(\tau_{rx}\hat{e}_x\otimes\hat{e}_x+\tau_{rr}\hat{e}_x\otimes\hat{e}_r+\tau_{r\varphi}\hat{e}_x\otimes\hat{e}_\varphi\right)
\end{equation}
Make other term via transpose.
\begin{equation}
	\begin{split}
		0 &= \frac{\partial u_x}{\partial r}\left[2\tau_{rx}\hat{e}_x\otimes\hat{e}_x+\tau_{rr}\left(\hat{e}_x\otimes\hat{e}_r+\hat{e}_r\otimes\hat{e}_x\right)+\tau_{r\varphi}\hat{e}_x\otimes\hat{e}_\varphi + \tau_{r\varphi}\hat{e}_\varphi\otimes\hat{e}_x\right]\\
		&- \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} + \frac{\eta_\text{p}}{\lambda}\frac{\partial u_x}{\partial r}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
	\end{split}
\end{equation}
If $\hat{e}_i\otimes\hat{e}_j$ does not appear in this equation $\tau_{ij} = 0$.
\begin{equation}
	0 = \frac{\partial u_x}{\partial r}\left[2\tau_{rx}\hat{e}_x\otimes\hat{e}_x\right]- \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} + \frac{\eta_\text{p}}{\lambda}\frac{\partial u_x}{\partial r}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
\end{equation}
If $\ma{\tau}$ is symmetric $\tau_{rx} = \tau_{xr}$.
\begin{equation}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} = \dot{\gamma}\left[2\tau_{xr}\hat{e}_x\otimes\hat{e}_x\right] + \frac{\eta_\text{p}}{\lambda}\dot{\gamma}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
\end{equation}
Separate by components.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xr} \\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xr} &= \eta_\text{p}\dot{\gamma}
\end{align}
The trace works out to:
\begin{equation}
	\Tr\ma{\tau} = \tau_{xx} + \tau_{rr} + \tau_{\varphi\varphi} = \tau_{xx}
\end{equation}
TODO: $N_1$? Defined along shear axis or always yy? ends up as $\tau_{xx}$ either way

}

\subsection{Channel flow}
For Poiseuille flow, we start with the Cauchy momentum equation (neglecting gravity), which reads as follows.
\begin{equation}
	\rho\frac{\Dif\vec{u}}{\Dif t} = -\nabla p + \nabla\cdot\sigma
\end{equation}
In steady state for this geometry, once integrated along $y$, this reduces to the following.
\begin{equation}
	\sigma_{xy} = \frac{\partial p}{\partial x}y
\end{equation}
The total stress experienced by the fluid is the one described by the constitutive equation plus the solvent contribution and thus the following holds.
\begin{equation}
	\tau_{xy} = \partial_xpy - \eta_\text{s}\dot{\gamma}
\end{equation}
Dividing equation~(\ref{eq:PTTDerivation1_2D}) by equation~(\ref{eq:PTTDerivation2_2D}) leads to the following relation.
\begin{equation}
	\tau_{xx} = \frac{2\lambda}{\eta_\text{p}}\tau_{xy}^2 =  \frac{2\lambda}{\eta_\text{p}}\left(\partial_xpy - \eta_\text{s}\dot{\gamma}\right)^2
\end{equation}
Inserting this back into equation~(\ref{eq:PTTDerivation2_2D}) yields the following.
\begin{align}
	\dot{\gamma} &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xpy - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xpy-\eta_\text{s}\dot{\gamma}\right)
	\intertext{This cannot be integrated as easily as without $\eta_\text{s}$, and is only solvable semi-analytically.}
	1 &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xpy - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{y}{\dot{\gamma}}-\eta_\text{s}\right) \\
	0 &= \frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xpy - \eta_\text{s}\dot{\gamma}\right]^2 - \ln\eta_\text{p} + \ln\left(\partial_xp\frac{y}{\dot{\gamma}}-\eta_\text{s}\right) \\
	0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{y}{\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{y}{\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right) \\
	0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{y}{\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{y}{\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right)
	\intertext{This has the following form}
	0 &= ab^2 + \ln b
\end{align}
This can be solved using the Lambert $W$ function as follows.
\begin{align}
	b &= \sqrt{\frac{W_0\left(2a\right)}{2a}}
	\intertext{Resolving for $y$ yields the following.}
	y &= \frac{\eta_\text{s}}{\partial_xp}\dot{\gamma} + \sgn\left(\dot{\gamma}\right)\frac{\eta_\text{p}}{2\lambda\sqrt{\epsilon}\partial_xp}\sqrt{W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)}
\end{align}
We notice, that this is the solution for a poiseuille flow with viscosity $\eta_\text{s}$ expanded by a term that applies a nonuniform scaling of the profile  along the $y$-axis towards higher absolute values. We simplify again using $c < 0$ and $d > 0$.
\begin{equation}
	y = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}\label{eq:semiAnalyticGD_2D}
\end{equation}\todo{sign and rest of explainer at the end?}
Finally this function needs to be inverted. This needs to be done numerically. We calculate a range of $\dot{\gamma}$ from $0$ to $\frac{R}{c}$ with a small step interval. Here $R$ is the radius of the channel or pipe. This is guaranteed to contain all relevant radii. We flip the axis and integrate numerically using \pyth{scipy.integrate.cumulative_trapezoid} \cite{scipy}. This is very accurate as we conveniently picked the spacing of our now $x$-axis in a way, that the $y$-axis ($\dot{\gamma}$) varies little for each step. From the recovered curve $u$ is calculated at each required $r$ through linear interpolation.

\subsection{Pipe flow}

For Poiseuille flow, we start with the Cauchy momentum equation (neglecting gravity), which reads as follows.

@@

		r = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}\label{eq:semiAnalyticGD_3D}
